---
title: "Introduction to ggplot2, rmarkdown; Advanced Data Manipulations with dplyr"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
description: >
  This week we will introduce the ggplot2 plotting framework, as well as the
  reproducible research framework `rmarkdown`. Time allowing, we'll also explore
  the join functionality of `dplyr`.
---

```{r setup, include=FALSE}
library(learnr)
library(stringr)
library(tidyverse)
library(tibble)
library(ggplot2)
library(nycflights13)
```

## Goals

This week, we have three goals: 

- Introduction of the `rmarkdown` framework
- First steps with the `ggplot2` plotting library
- Diving deeper into `dplyr`

## RMarkdown

We begin with the `rmarkdown` framework. 

Open the `demo.Rmd` file in your `RStudio` window. Press the `knit` button at the
top of the file editor to confirm that you have your environment set up properly. 
(There shouldn't be an issue if you could render this file, but it's worth checking)

Once that pops up, close it for now and come back here.

### Documents as Code

`Markdown` (`md`) is a document markup language (get the pun?) originally designed
to make it easy to write web-pages. Because it was designed for websites, markdown
is particularly well-suited for making `html` documents like this one, though
other output formats  (*e.g.* PDFs) are possible. 

Before exploring Markdown, let's pause and define what we mean by a _markup_ language.
Document formats (and associated software) can be divided into two groups:

- `WYSIWYG` (What you see is what you get): In these formats, you directly work
  with the 'rendered' version of a document, and no post-processing is needed.
  Microsoft Word is a classic example: what you see on your screen is what it
  looks like on the page (assuming no printer errors).
- `WYSIWYM` (What you see is what you mean): In these formats, you essentially
  write *code* that defines both the content and formatting of a document. 
  These formats can be complex (various XML based formats) or relatively simple
  like Markdown. 
  
Why would we ever prefer WYSIWYM styles over alternatives? 

- Ease of editing: because WYSIWYM formats are not tied to specific software, 
  they can be edited anywhere.
- Flexibility and Extensibility: When working in MS Word, you're basically only
  able to create 'Word-type' documents. If you want to swap to a presentation, 
  you have to exit Word and open Powerpoint. In WYSIWYM world, the same document
  can be easily transformed from one format to another.
- Separation of content from form. As we will see, WYSIWYM formats mark text as "important" 
  or a "section header" (or similar), but leave the choice of what that means to 
  software. If you want to later change from bold to italics for emphasis, you don't 
  need to re-edit the entire document. 
  
Additionally, because WYSIWYM documents are plain text files, they can be used to
create accessible documents more easily than WYSIWYG formats. 

## First Steps with Markdown

Markdown can be divided into four major formatting tools: 

- Section Headers
- Emphasis Formatting
- Paragraph Structure
- Web Content (e.g., links, embedded code, video *etc.*)

Section headers are denoted by lines beginning with one or more `#` symbols.
A single `#` represents a major title, with `##` representing sections, `###` subsections, *etc*.
Most Markdown implementation stops at `####` but there is no reason why this couldn't be
continued. 

Emphasis formatting consists of one or more `*` characters surrounding the text to be
emphasized: 

- *one star*
- **two stars**
- ***three stars***

Paragraphs are separated by a single blank line. New lines without space between
them have no semantic meaning to markdown. They are just used to keep your code
from getting too long (80 characters per line is a typical recommendation.)

To create bulleted list, start one or more lines with a `-` character. To create numbered
lists, start one or more lines with `1.`. (You can actually use any numbers, including
repeats - Markdown is smart enough to know what you want)

### Markdown Exercises

Open `demo1.Rmd` and make the following changes: 

1. Add your name to the header block
2. Add a bullet list of your classes to the "About Me" section
3. Add a new section titled "My Projects" and add a numbered list of
   projects you have done for this and other classes. **Emphasize** each title.
   
Once you have done these, render (knit) the document and confirm that your changes
took place. 

## More Markdown

In addition to basic formatting, Markdown supports many "programmer-relevant" 
features, including code-style formatting `1+1=2`, [links](https://commonmark.org/help/tutorial), 
and inline code blocks

```{r, eval=FALSE}
library(dplyr)
library(nycflights13)
glimpse(flights)
```

Go to the [CommonMark Tutorial](https://commonmark.org/help/tutorial) to explore these
and more features of Markdown. 

### More Markdown Exercises

1. The St. Louis Blues are a jazz standard. Add a link to a YouTube performance of the
   blues (by anyone) into your Markdown document. 
 
2. Now turn the link into a embedded image / video. 
   
3. Use RStudio's built in templates to create a basic presentation using RMarkdown. 
   (Hint: File > New)

## But Wait - there's moRe!

Ok - so that's the basics of *Markdown*. What is *RMarkdown*? 

*RMarkdown* is the fusion of the computational tools of *R* with the 
formatting tools of *Markdown*. 

In short, *RMarkdown* adds a pre-processing hook to run *R* code before handing
your document over to the rendering engine. *RMarkdown* will find any *R* code in your document, 
run it and format the output for Markdown, and then pass the resulting pure-Markdown document
off to the rendering engine. 

Most of the time, we'll use code blocks like the one we showed above. These
are blocked out with triple backticks and a little `r` (see the source of this document
for many examples). Triple backticks indicate the end of a block. 

You might have seen a similar idea with *Jupyter/IPython Notebook Cells*. There
is a big conceptual difference however: 
- Notebooks are executed manually (by the data analyst) and cells can be run
  in any order (as long as that doesn't produce an error)
- Markdown documents are run automatically "in order", so you have to define 
  a variable earlier ("higher up") than when you want to use it.
  
### RMarkdown Exercises

1. Take `demo.Rmd` and modify the code block to print the 10 most delayed flights.

2. Add a code block that prints the numbers 1 to 10 in a loop. How does RMarkdown format the output?

3. You can silence certain code blocks by adding `echo=FALSE` (to hide the code) 
   or `include=FALSE` (to hide the output) to the block options. 
   Add a code block to print "Hello, I love R Markdown" five times, but hide the code
   and only show the output. 
   
4. You can also include the results of small snippets of *R* code using inline code: 
   *e.g*. two times pi is approximately equal to `{r} 2*pi`. Add an inline block
   to say how many rows are in the flights data frame. (Hint: you will want the `NROW` function.)

   
## Time to ggplot

*R* has (at least) three popular sets of plotting functionality. We will focus on
the **ggplot2** library in this course. 

The name **ggplot2** comes from the term "Grammar of Graphics." (It's the second version
of software implementing these ideas). The grammar of graphics refers to the idea
that you can build plots systematically from complex structures of simple elements.

These include: 

- `geom`s: the actual "thing" drawn on the screen
- `aes`thetics: a mapping from some feature of the data to some graphical characteristic
   (*e.g.* x-axis location or color)

Let's dig into `ggplot2`. We'll begin with the `penguins` dataset. 

```{r}
library(palmerpenguins)
penguins
```

This is a table of 344 penguins (rows) from the Palmer antarctic research base. Scientists recorded
the following 8 features (columns)

- species
- island
- bill_length_mm
- bill_depth_mm
- flipper_length_mm
- body_mass_g
- sex
- year

Run the `?penguins` command to get information about each of these features. 

```{r penguinhelp, exercise = TRUE}
?penguins
```

We can plot this data using the `ggplot` command: 

```{r gg1, exercise = TRUE}
library(palmerpenguins)
ggplot(penguins)
```


Awesome, no?

### Adding more to our plots

In order to get a useful plot, we need to add some of those geoms and aesthetics we
discussed earlier. Let's first see how flipper and bill size are related by placing
them on the `x` and `y` aesthetics. 

```{r}
library(palmerpenguins)

ggplot(penguins, aes(x=bill_length_mm, flipper_length_mm))
```

This sort of starts to look like a plot, but we don't have any data points shown. 

To actually show plots, we need a "geom" (geometric element) that will actually
show our data. Let's start with a  simple scatter plot: 

```{r}
library(palmerpenguins)

ggplot(penguins, aes(x=bill_length_mm, flipper_length_mm)) + geom_point()
```

Note that this is different than our `dplyr` analysis - we _add_ layers to the plot
using the `+` operator; we don't _pipe_ our data using the `|>` operator.

This is nice, but can we do better? How does this relationship vary by species? 
Modify the previous examples to map the `species` variable to the `color` aesthetic. 

```{r gg2, exercise=TRUE}
library(palmerpenguins)

ggplot(penguins, aes(x=bill_length_mm, flipper_length_mm)) + geom_point()
```

### Making Nicer Plots

Now that we have a basic plot, let's focus on some ways of making it more attractive: 
- axis labels
- plot label
- better colors
- adding trend lines

We can add labels to the x, y axes with the `xlab` and `ylab` functions. Change
the x-axis label to "Bill Length (mm)" and the y-axis label to "Flipper Length (mm)". 

```{r gg3, exercise=TRUE}
library(palmerpenguins)

ggplot(penguins, aes(x=bill_length_mm, flipper_length_mm, color=species)) + geom_point()
```

Remember - units are important! Also - while variable names are useful, you typicaly want
to "English" them up when making actual plots for sharing.

Next, let's clean up the colors. `ggplot2` comes with many built in palettes (sets of colors), 
but I really like one from the `ggsci` package. 

```{r}
library(ggsci)
```

The `ggsci` package provides the `JCO` color scale. We can add it to our penguin plot

```{r}
ggplot(penguins, aes(x=bill_length_mm, flipper_length_mm, color=species)) + geom_point() + scale_color_jco()
```

Here we've _added_ a new scale to the plot, overriding the default. 

`ggplot2` also uses a greyscale theme which is easy to look at on a screen. 

Try out some of the [other color themes](https://ggplot2.tidyverse.org/reference/ggtheme.html) 
from the `ggplot2` documentation. 

```{r gg4, exercise=TRUE}
library(palmerpenguins)
library(ggsci)
ggplot(penguins, aes(x=bill_length_mm, flipper_length_mm, color=species)) + 
    geom_point() + scale_color_jco() + 
    theme_grey() ## CHANGE ME! theme_grey() is the default
```

Finally, add a smooth trend line using the `geom_smooth` function. 

```{r gg5, exercise=TRUE}
library(palmerpenguins)

ggplot(penguins, aes(x=bill_length_mm, flipper_length_mm, color=species)) + 
    geom_point() + scale_color_jco() + 
    theme_grey() ## CHANGE ME! theme_grey() is the default
```

Now, fit a linear trend by passing the argument `method='lm'` to `geom_smooth`. 

To finish this plot, improve the label of the color legend by passing a `name` argument
to `scale_color_jco` and an overall plot title to `ggtitle` to make a publication
ready plot.

```{r gg6, exercise=TRUE}
library(palmerpenguins)
ggplot(penguins)
```

### ggplot Exercises

Make plots to demonstrate the following: 

- A histogram of bill depth
- Species specific box plots of body mass
- Violin plots comparing body mass by sex
- Violin plots comparing body mass by sex for different species: use the `facet_grid` function

The [ggplot2 docs](https://ggplot2.tidyverse.org) will be helpful. 

## Happy Leap Day

***Real Life is for March!***

![](https://www.youtube.com/watch?v=Bc9-VVYseH0)
